Stressing Bank Profitability for Interest Rate Risk$^*$

### 2013-07-09 LATEX/beg.draft.2013.tex
updated draft

\begin{abstract}
 Regular bank stress testing is one of the three complementary reforms that has been made to U.S. capital regulation in response to the shortcomings of the pre-crisis regime.
 The form of stress testing mandated by the Dodd-Frank Act for the Federal Reserve to undertake is one based around a small number of scenarios.  This approach involves specifying macroeconomic and financial scenarios that represents stressful conditions over the time horizon of the stress test and then projecting, for each bank holding company in the stress test, its losses, income, and path of capital \textit{given that scenario}. In this paper, we focus on the path of income. In particular, we consider the ability of empirical models to translate developments in interest rates into bank net interest margins (NIMs).

@@

\vspace{2cm}
\noindent $^*$ The views
expressed in this paper are solely the responsibility of the
authors and should not be interpreted as reflecting the views of
the Board of Governors of the Federal Reserve System or of any
other person associated with the Federal Reserve System.
Contact Information: valentin.bolotnyy@frb.gov; rochelle.m.edge@frb.gov; luca.guerrieri@frb.gov.

\end{abstract}
\renewcommand{\baselinestretch}{1.5}
\thispagestyle{empty}

@@

Scenario based stress tests are predicated on the ability to produce reliable conditional forecasts for key measures of profitability such as net interest margins. This kind of stress test program is not limited to the U.S. banking system. Stress testing of banks based on scenarios was undertaken during the Scandinavian banking crisis to shore up confidence in the ability of banks to withstand the crisis. Most extensively, the European Banking Authority conducted stress tests based on scenario analysis throughout the European Union in 2009, 2010, and 2011. The results of such programs can only be effective in restoring confidence from the public in the midst of a crisis to the extent that the public can be convinced of the ability of regulators to assess the consequences of their scenarios on banks. If forecast methods at the frontier of efficiency have disentangling the implications of baseline and severe stress scenarios, the ability of stress test programs to avoid panic is called in question.

@@

We construct forecasts from a wide array of methods taken from the forecasting literature. We used data with quarterly frequency and consider a forecast horizon up to ten quarters long. We are interested in forecasts for NIMs conditional on the variables in scenario $I$, which we denote $NIM_{t+s/I}$. Here are the methods considered.

\vspace{-0.1in}
\noindent \textbf{1. Simple forecast combination.} We regress NIMs on each yield $r(\tau )_{t-1}$ separately: \
\[
NIM_{t}=c_{\tau }+\rho _{\tau }NIM_{t-1}+\sum_{j=1}^4\gamma _{\tau,j}r(\tau)_{t-j}+{\eta
_{\tau ,t}}
\]
Each regression is used to produce an $s$ step-ahead forecast of NIMs conditional on yields on treasuries r(\tau) in scenario $I$ with maturity $\tau$ observed through period $t+s-1.$ This conditional forecast is denoted by $NIM_{\tau,t+s/I}.$ The simple forecast comnination is then given by:
\[
NIM_{t+s/I}=\sum_{\tau }\frac{NIM_{\tau ,t+s/I}}{N},
\]
where $N$ is the number of maturities considered.

\noindent \textbf{2. A dynamic factor model for NIMs.} Using the state space representation, the state equations of our restricted dynamic factor model is given by
\[
\left(
\begin{array}{c}
{L_{t+1}-\mu _{L}} \\
{S_{t+1}-\mu _{S}} \\
{C_{t+1}-\mu _{C}} \\
\widehat{NIM}_{t+1}-\mu _{\widehat{NIM}}%
\end{array}%
\right) {\normalsize \ =\ }\left(
\begin{array}{cccc}
a_{11} & a_{12} & a_{13} & 0 \\
a_{21} & a_{22} & a_{23} & 0 \\
a_{31} & a_{32} & a_{33} & 0 \\
a_{41} & a_{42} & a_{43} & a_{44}%
\end{array}%
\right) {\normalsize \ }\left(
\begin{array}{c}
{L_{t}-\mu _{L}} \\
{S_{t}-\mu _{S}} \\
{C_{t}-\mu _{C}} \\
\widehat{NIM}_{t}-\mu _{\widehat{NIM}}%
\end{array}%
\right) {\normalsize +}\left(
\begin{array}{c}
{\eta _{Lt}} \\
{\eta _{St}} \\
{\eta _{Ct}} \\
{\eta _{NIMt}}%
\end{array}%
{\normalsize \ }\right) ,
\]
where $L_{t},S_{t},$ and $C_{t}$ are level slope and curvature factors for the yield curve, respectively, and $\widehat{NIM}_{t}$ represent a measure of NIMs cleaned of observation error. The term $\mu _{x}$ denotes the mean of $x$. The terms ${\eta _{Lt}}$, ${\eta _{St}}$, ${\eta _{Ct}}$, ${\eta_{NIMt}}$ denote independently distributed innovations with joint covariance matrix $\Omega $.

The observation equations, which relate a set of $N$ yields (in our case, 12) and observed NIMs can be split into two groups. In the first group, equations for yields on treasuries of maturity $\tau $ take the form:
\[
r(\tau )=\left(
\begin{array}{cccc}
1 & \frac{1-e^{-\lambda \tau }}{\lambda \tau } & \frac{1-e^{-\lambda \tau }}{%
\lambda \tau }-e^{-\lambda \tau } & 0%
\end{array}%
\right) \left(
\begin{array}{c}
{L_{t}-\mu _{L}} \\
{S_{t}-\mu _{S}} \\
{C_{t}-\mu _{C}} \\
\widehat{NIM}_{t}-\mu _{\widehat{NIM}}%
\end{array}%
\right) +e(\tau )_{t},
\]%
where $e(\tau )_{t}$ is i.i.d. measurement error assumed to be uncorrelated
across observations. \bigskip In the second group, the equation for observed
NIMs takes the form%
\[
NIM_{t}=\left(
\begin{array}{cccc}
0 & 0 & 0 & 1%
\end{array}%
\right) \left(
\begin{array}{c}
{L_{t}-\mu _{L}} \\
{S_{t}-\mu _{S}} \\
{C_{t}-\mu _{C}} \\
\widehat{NIM}_{t}-\mu _{\widehat{NIM}}%
\end{array}%
\right) +e(NIM)_{t},
\]
where $e(NIM)_{t}$ is i.i.d. measurement error assumed to be uncorrelated across observations.

Out-of-sample conditional forecasts for NIMs from this model, are obtained using the Kalman filter.

\noindent \textbf{3. Dynamic factor model with second-step regression for NIMs.} The dynamic factor model we consider takes the form
\[
\left(
\begin{array}{c}
{L_{t+1}-\mu _{L}} \\
{S_{t+1}-\mu _{S}} \\
{C_{t+1}-\mu _{C}}%
\end{array}%
\right) {\normalsize \ =\ }\left(
\begin{array}{ccc}
a_{11} & a_{12} & a_{13} \\
a_{21} & a_{22} & a_{23} \\
a_{31} & a_{32} & a_{33}%
\end{array}%
\right) {\normalsize \ }\left(
\begin{array}{c}
{L_{t}-\mu _{L}} \\
{S_{t}-\mu _{S}} \\
{C_{t}-\mu _{C}}%
\end{array}%
\right) {\normalsize +}\left(
\begin{array}{c}
{\eta _{Lt}} \\
{\eta _{St}} \\
{\eta _{Ct}}%
\end{array}%
\right)
\]

\[
r(\tau )=\left(
\begin{array}{ccc}
1 & \frac{1-e^{-\lambda \tau }}{\lambda \tau } & \frac{1-e^{-\lambda \tau }}{%
\lambda \tau }-e^{-\lambda \tau }%
\end{array}%
\right) \left(
\begin{array}{c}
{L_{t}-\mu _{L}} \\
{S_{t}-\mu _{S}} \\
{C_{t}-\mu _{C}} \\
\widehat{NIM}_{t}-\mu _{\widehat{NIM}}%
\end{array}%
\right) +e(\tau )_{t},
\]

We obtain smoothed estimates of the factors $L_{t},S_{t},$ and $C_{t}$ conditional on information up until period $T.$ We estimate two alternative second-step forecasting equations:

\noindent \textbf{Alternative a. Multivariate} (all yield curve factors included in one NIM equation)
\[
NIM_{t}=c+\rho NIM_{t-1}+\gamma _{L}L_{t-1}+\gamma _{S}S_{t-1}+\gamma
_{C}C_{t-1}
\]

\noindent \textbf{Alternative b. Forecast combination} (each factor included in a separate equation and forecasts from each equation combined)
\[
NIM_{t}=c_{i}+\rho _{i}NIM_{t-1}+\gamma _{i}F_{i,t-1}+{\eta _{i,t},}
\]%
where $F_{i}\in \left\{ L,S,C\right\} .$ Forecasts $NIM_{i,t+s/t}$ from each separate regression are then aggregated as:

\[
NIM_{t+s/t}=\sum_{i}\frac{NIM_{i,t+s/t}}{3},
\]

\noindent \textbf{4. Forecast combination with observed factors.} Define $\left\{L,S,C \right\} ,$ the observed factors as in subsection 3.1 above.
\[
NIM_{t}=c_{i}+\rho _{i}NIM_{t-1}+\gamma _{i}O_{i,t-1}+{\eta _{i,t},}
\]%
where $O_{i}\in \left\{ L,S,C\right\} .$ Forecasts  from each separate regression, $NIM_{i,t+s/t},$are then aggregated as:

\noindent \textbf{5. VAR with observed factors.} Define level factor to coincide with the 10-year yield; define the slope factor as the difference between the 10-year and the 3-month yields; define the curvatures factor to be 2 times the 2-year yield minus the the sum of the 3-month and 10-year yields. The VAR of order four includes the three observed factors defined as above and NIMs. Forecasts conditional on the factors are obtained using the Kalman filter.
\[
NIM_{t+s/t}=\sum_{i}\frac{NIM_{i,t+s/t}}{3},
\]

\noindent \textbf{6. No-change forecast.} According to this model $NIM_{t+s/t}=NIM_{t}${\normalsize . }

>>>>

\begin{abstract}
Regular bank stress testing is one of the three complementary reforms that has been made to U.S. capital regulation in response to the shortcomings of the pre-crisis regime.
The form of stress testing mandated by the Dodd-Frank Act for the Federal Reserve to undertake is one based around a small number of scenarios.  This approach involves specifying macroeconomic and financial scenarios that represents stressful conditions over the time horizon of the stress test and then projecting, for each bank holding company in the stress test, its losses, income, and path of capital \textit{given that scenario}. In this paper, we focus on the path of income. In particular, we consider the ability of empirical models to translate developments in interest rates into bank net interest margins (NIMs).
\end{abstract}

@@

\vspace{3cm} \renewcommand{\baselinestretch}{1} \normalsize
\noindent $^*$ The views
expressed in this paper are solely the responsibility of the
authors and should not be interpreted as reflecting the views of
the Board of Governors of the Federal Reserve System or of any
other person associated with the Federal Reserve System.
Contact Information: valentin.bolotnyy@frb.gov; rochelle.m.edge@frb.gov; luca.guerrieri@frb.gov.

\renewcommand{\baselinestretch}{1.5} \normalsize
\thispagestyle{empty}

@@

Scenario based stress tests are predicated on the ability to produce reliable conditional forecasts for key measures of profitability such as net interest margins. This kind of stress test program is not limited to the U.S. banking system. Stress testing of banks based on scenarios was undertaken during the Scandinavian banking crisis to shore up confidence in the ability of banks to withstand the crisis. Most extensively, the European Banking Authority conducted stress tests based on scenario analysis throughout the European Union in 2009, 2010, and 2011. The results of such programs can only be effective in restoring confidence from the public in the midst of a crisis to the extent that the public can be convinced of the ability of regulators to assess the consequences of their scenarios on banks. If forecast methods at the frontier of efficiency have trouble disentangling the implications of baseline and severe stress scenarios, the ability of stress test programs to avoid panic is called in question.

@@

We construct forecasts from a wide array of standard methods taken from the forecasting literature. We use data with quarterly frequency and consider a forecast horizon up to ten quarters long. We are interested in forecasts for NIMs conditional on the variables in scenario $I$, which we denote $NIM_{t+s/I}$. Here are the methods considered:

\vspace{-0.1in}
\noindent \textbf{1. Equal-weighted forecast combination.} We regress NIMs on each yield $r(\tau )_{t-1}$ separately: \
\[
NIM_{t}=c_{\tau }+\rho _{\tau }NIM_{t-1}+\sum_{j=1}^4\gamma _{\tau,j}r(\tau)_{t-j}+{\eta
_{\tau ,t}}
\]
Each regression is used to produce an $s$ step-ahead forecast of NIMs conditional on yields on treasuries $r(\tau)$ in scenario $I$ with maturity $\tau$ observed through period $t+s-1.$ This conditional forecast is denoted by $NIM_{\tau,t+s/I}.$ The equal-weighted forecast combination is then given by:
\[
NIM_{t+s/I}=\sum_{\tau }\frac{NIM_{\tau ,t+s/I}}{N},
\]
where $N$ is the number of maturities considered.\footnote{Experimentation with alternative lag structures led to no improvement in the forecast results reported below.}

\noindent \textbf{2. A dynamic factor model for NIMs.} We include NIMs in the dynamic factor model for the yield curve developed by Diebold and Li (2006). Using the state space representation, the state equations of our dynamic factor model is given by:
\[
\left(
\begin{array}{c}
{L_{t+1}-\mu _{L}} \\
{S_{t+1}-\mu _{S}} \\
{C_{t+1}-\mu _{C}} \\
\widehat{NIM}_{t+1}-\mu _{\widehat{NIM}}%
\end{array}%
\right) {\normalsize \ =\ }\left(
\begin{array}{cccc}
a_{11} & a_{12} & a_{13} & 0 \\
a_{21} & a_{22} & a_{23} & 0 \\
a_{31} & a_{32} & a_{33} & 0 \\
a_{41} & a_{42} & a_{43} & a_{44}%
\end{array}%
\right) {\normalsize \ }\left(
\begin{array}{c}
{L_{t}-\mu _{L}} \\
{S_{t}-\mu _{S}} \\
{C_{t}-\mu _{C}} \\
\widehat{NIM}_{t}-\mu _{\widehat{NIM}}%
\end{array}%
\right) {\normalsize +}\left(
\begin{array}{c}
{\eta _{Lt}} \\
{\eta _{St}} \\
{\eta _{Ct}} \\
{\eta _{NIMt}}%
\end{array}%
{\normalsize \ }\right) ,
\]
where $L_{t},S_{t},$ and $C_{t}$ are level slope and curvature factors for the yield curve, respectively, and $\widehat{NIM}_{t}$ represent a measure of NIMs cleaned of observation error. The term $\mu _{x}$ denotes the mean of $x$. The terms ${\eta _{Lt}}$, ${\eta _{St}}$, ${\eta _{Ct}}$, ${\eta_{NIMt}}$ denote independently distributed innovations with joint covariance matrix $\Omega $.

The observation equations, which relate a set of $N$ yields (in our case, 12) and observed NIMs can be split into two groups. In the first group, equations for yields on treasuries of maturity $\tau $ take the form:
\[
r(\tau )=\left(
\begin{array}{cccc}
1 & \frac{1-e^{-\lambda \tau }}{\lambda \tau } & \frac{1-e^{-\lambda \tau }}{%
\lambda \tau }-e^{-\lambda \tau } & 0%
\end{array}%
\right) \left(
\begin{array}{c}
{L_{t}-\mu _{L}} \\
{S_{t}-\mu _{S}} \\
{C_{t}-\mu _{C}} \\
\widehat{NIM}_{t}-\mu _{\widehat{NIM}}%
\end{array}%
\right) +e(\tau )_{t},
\]%
where $e(\tau )_{t}$ is i.i.d. measurement error assumed to be uncorrelated
across observations. \bigskip In the second group, the equation for observed
NIMs takes the form%
\[
NIM_{t}=\left(
\begin{array}{cccc}
0 & 0 & 0 & 1%
\end{array}%
\right) \left(
\begin{array}{c}
{L_{t}-\mu _{L}} \\
{S_{t}-\mu _{S}} \\
{C_{t}-\mu _{C}} \\
\widehat{NIM}_{t}-\mu _{\widehat{NIM}}%
\end{array}%
\right) +e(NIM)_{t},
\]
where $e(NIM)_{t}$ is i.i.d. measurement error assumed to be uncorrelated across observations. Out-of-sample conditional forecasts for NIMs from this model, are obtained using the Kalman filter. We focus on the filtered estimates of NIMs given a path for all yields included in the model.

\noindent \textbf{3. Dynamic factor model with second-step regression for NIMs.} This model extracts level, slope, and curvature factors for the yield curve following Diebold and Li (2006). The factors are used to forecast NIMs in a separate second-step regression. The dynamic factor model we consider takes the form
\[
\left(
\begin{array}{c}
{L_{t+1}-\mu _{L}} \\
{S_{t+1}-\mu _{S}} \\
{C_{t+1}-\mu _{C}}%
\end{array}%
\right) {\normalsize \ =\ }\left(
\begin{array}{ccc}
a_{11} & a_{12} & a_{13} \\
a_{21} & a_{22} & a_{23} \\
a_{31} & a_{32} & a_{33}%
\end{array}%
\right) {\normalsize \ }\left(
\begin{array}{c}
{L_{t}-\mu _{L}} \\
{S_{t}-\mu _{S}} \\
{C_{t}-\mu _{C}}%
\end{array}%
\right) {\normalsize +}\left(
\begin{array}{c}
{\eta _{Lt}} \\
{\eta _{St}} \\
{\eta _{Ct}}%
\end{array}%
\right)
\]

\[
r(\tau )=\left(
\begin{array}{ccc}
1 & \frac{1-e^{-\lambda \tau }}{\lambda \tau } & \frac{1-e^{-\lambda \tau }}{%
\lambda \tau }-e^{-\lambda \tau }%
\end{array}%
\right) \left(
\begin{array}{c}
{L_{t}-\mu _{L}} \\
{S_{t}-\mu _{S}} \\
{C_{t}-\mu _{C}} 
\end{array}%
\right) +e(\tau )_{t},
\]
We obtain smoothed estimates of the factors $L_{t},S_{t},$ and $C_{t}$ conditional on information up until period $T.$ We estimate two alternative second-step forecasting equations:

\noindent \textbf{Alternative a. Multivariate} (all yield curve factors included in one NIM equation)
\[
NIM_{t}=c+\rho NIM_{t-1}+\gamma _{L}L_{t-1}+\gamma _{S}S_{t-1}+\gamma
_{C}C_{t-1}
\]
We use this equation to produce recursive forecasts for NIMs conditional on the the smoothed factors.

\noindent \textbf{Alternative b. Forecast combination} (each factor included in a separate equation and forecasts from each equation combined)
\[
NIM_{t}=c_{i}+\rho _{i}NIM_{t-1}+\gamma _{i}F_{i,t-1}+{\eta _{i,t},}
\]%
where $F_{i}\in \left\{ L,S,C\right\}.$ Recursive forecasts $NIM_{i,t+s/I}$ from each separate regression are then aggregated as:
\[
NIM_{t+s/I}=\sum_{i}\frac{NIM_{i,t+s/I}}{3},
\]

\textbf{4. Principal component regression}
Let $X$ a $T \times n$ matrix of $n$ Treasury yields.  Using the spectral decomposition, one obtains $X'X = VDV'$. Let $V_i$ be the normalized eigenvector corresponding to $i^{th}$ largest eigenvalue.
The $i^{th}$ principal component is defined as $PC_i = XV_i$. We estimate
\begin{equation*}
NIM_t = c + \rho NIM_{t-1} + \sum_{i=1}^3 \gamma_i PC_{i,t-1} + e_t.
\end{equation*}
using ordinary least squares and we construct recursive forecasts for net interest margins conditional on the principal components.

\begin{frame}

\noindent \textbf{5. Forecast combination with observed factors.} Define level factor $L$ to coincide with an average of the 3-month, 2-year and 10-year Treasury yields; define the slope factor as the difference between the 10-year and the 3-month yields; define the curvatures factor to be 2 times the 2-year yield minus the the sum of the 3-month and 10-year yields. Define $\left\{L,S,C \right\} ,$ the observed factors as in subsection 3.1 above.
\[
NIM_{t}=c_{i}+\rho _{i}NIM_{t-1}+\gamma _{i}O_{i,t-1}+{\eta _{i,t},}
\]%
where $O_{i}\in \left\{ L,S,C\right\} .$ Forecasts  from each separate regression, $NIM_{i,t+s/t},$are then aggregated as:

\noindent \textbf{6. VAR with observed factors.}  The VAR of order four includes the three observed factors defined as above and NIMs. Forecasts conditional on the factors are obtained using the Kalman filter.

\noindent \textbf{7. No-change forecast.} According to this model $NIM_{t+s/t}=NIM_{t}${\normalsize . }
